---
title: "EPID674: FINAL PROJECT"
author: 
  - "Kurt M. Kunz, MD"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

# Final Project Tasks

* For your final project, you will choose from two data sets to study:

    - (1) an OTU table from a study of respiratory tract microbiome (OP = oropharyngeal, ET = endotracheal), which is described here: https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-016-0151-8   
    - (2) an OTU table from a study of mouse cecal and stool from different mouse vendors 

* There are four tasks to complete for the final project:  

1. Alpha diversity:  
    - A. Calculate Shannon diversity for all specimens included in your study of choice  
    - B. Plot the Shannon diversity with separate colors for specimen type ("OP" vs "ET" specimens for data set #1; or vendor name for data set #2).  
    - HINT: use the long form of the data and group by specimen  

2. Develop a linear model for the Shannon diversity:  
    - for data set #1: Shannon diversity ~ ET/OP  
    - for data set #2: Shannon diversity ~ vendor, or Shannon diversity ~ specimen_type
    - HINT: use the output from above

3. Beta diversity:
    - A. Use the `vegdist` function to calculate Jaccard distances between specimens
    - B. Use the `adonis` function to assess the effect of ET/OP (data set #1) or Cecal/Stool (data set #2)
    - HINT: use the matrix form of the data... and don't forget to transpose!

4. Dirichlet Multinomial Mixture model:
    - A. Apply a DMM model to your chosen data set -- how many community types are recognized
    - B. Use the "fitted" function to assign categorical mixture type to each specimen
    - HINT: use the matrix form of the data... and don't forget to transpose... then use the distance matrix as the dependent variable of the PERMANOVA model


## package and environment settings

```{r load_pkgs, echo=TRUE, eval=TRUE}

# remember to install packages the first time!
#' install.packages("tidyverse")
#' install.packages("vegan")
#' install.packages("BiocManager")
#' BiocManager::install("DirichletMultinomial")

library(tidyverse)
library(vegan)
library(ape)
library(DirichletMultinomial)
set.seed(16)

# ggplot2 settings
theme_set(theme_bw())
```

## Import data

```{r load_otus, echo=TRUE, eval=TRUE, fig.align='center'}

icu_long <- read_csv(file = "./final_project_data/icu_small_long.csv.gz") %>%
  select(-specimen_source, specimen_source)
icu_mat <- read_rds(path = "./final_project_data/icu_small_mat.rds")

# mouse_long <- read_csv(file = "./final_project_data/mouse_small_long.csv.gz")
# mouse_mat <- read_rds(path = "./final_project_data/mouse_small_mat.rds")

```

## __Task 1: Alpha Diversity__

Among all specimens combined, alpha diversity as estimated by Shannon index (base 2) was 8.22. The Shannon diversity was 6.75 among specimens obtained from endotracheal tube and 7.59 among specimens obtained from oropharyngeal sampling.

```{r task1, echo=TRUE, eval=TRUE, fig.align='center'}

icu_long %>%
  summarise(
    specimen_shannon = diversity(x = read_count,
                                 index = "shannon",
                                 base = 2))

icu_long %>%
  group_by(specimen_source) %>%
  summarise(
    specimen_shannon = diversity(x = read_count,
                                 index = "shannon",
                                 base = 2))

icu_shannon <-
icu_long %>%
  group_by(specimen_id) %>%
  mutate(
    specimen_shannon = diversity(x = read_count,
                                 index = "shannon",
                                 base = 2),
    # # rarefied Shannon diversity
    # rare_count = as.vector(
    #   rrarefy(read_count, sample = 1000)),
    # rare_shannon = diversity(x = rare_count,
    #                          index = "shannon",
    #                          base = 2)
    
  ) %>%
  ungroup()

icu_shannon %>%
  ggplot(data = ., aes(x = specimen_source, y = specimen_shannon)) +
  geom_boxplot(aes(fill = specimen_source), show.legend = F) +
  labs(title = "Shannon Diversity by Specimen Source",
         x = "Specimen Source",
       y = "Shannon diversity (base 2)")

```

## Task 2: Linear Model


```{r task2, echo=TRUE, eval=TRUE, fig.align='center'}

mod <-
  lm(specimen_shannon ~ specimen_source,
     data = icu_shannon)

mod %>% summary()

ggplot(data = icu_shannon, 
       aes(x = specimen_source, 
           y = specimen_shannon)) +
  geom_point(aes(color = specimen_source), show.legend = F) +
  geom_smooth(method = "lm") +
  labs(title = "Shannon Diversity by Specimen Source",
       x = "Specimen Source",
       y = "Shannon Index (base 2)")

```


## Task 3: Beta Diversity

### 3a. Jaccard distance matrix with vegdist()

Here I calculate weighted Jaccard distances and save these to icu_dist. These are piped into matrix form in order to plot and compare between-group vs within-group distances.

```{r task3a, echo=TRUE, eval=TRUE, fig.align='center'}

# create jaccard weighted distance matrix
icu_dist <-
icu_mat %>% #<<
  t() %>% # transpose
  vegdist(x = .,
          method = "jaccard",
          binary = T) # use weighted distances

# create tibble from distance matrix, to plot distances
icu_dm <-
icu_dist %>% 
  as.matrix() %>%
  as_tibble(rownames = "specimen_1") %>%
  # convert to long format tibble
  pivot_longer(data = .,
               cols = starts_with("VAP"),
               names_to = "specimen_2",
               values_to = "weighted_jaccard") %>%
  # specimen 1 (vertical axis of matrix), merge specimen source
  left_join(x = ., 
            y = select(.data = icu_long,
                       specimen_id, specimen_source),
            by = c("specimen_1" = "specimen_id")) %>%
  distinct() %>%
  # specimen 2 (horizontal axis of matrix), merge specimen source
  left_join(x = .,
            y = select(.data = icu_long,
                       specimen_id, specimen_source),
            by = c("specimen_2" = "specimen_id")) %>%
  distinct()

# plot lumped distances
icu_dm %>%
  mutate(dist_cat =
           if_else(specimen_source.x != specimen_source.y,
                  "between",
                  specimen_source.x)) %>%
  ggplot(data = .,
         aes(x = weighted_jaccard)) +
  geom_histogram(
    aes(fill = dist_cat),
    alpha = 0.5,
    position = "identity") +
  scale_fill_brewer(palette = "Set1", direction = 1) +
  labs(title = "Between-Group vs Within-Group Distances",
      x = "Weighted Jaccard Distance")

```

### 3b. PERMANOVA with adonis()

```{r task3b, echo=TRUE, eval=TRUE, fig.align='center'}

# principle coordinates analysis, using jaccard distances from earlier estimate
# see lecture 8, slide 31
icu_pc <-
icu_dist %>%
  pcoa(D = .)

# create principle coordinates summary dataframe
icu_summary <- #save
icu_pc$vectors %>%
  as_tibble(rownames = "specimen_id") %>%
  select(specimen_id, Axis.1, Axis.2) %>%
  left_join(x = .,
            y = select(.data = icu_shannon,
                       specimen_id, specimen_source, specimen_shannon), 
            by = "specimen_id")

icu_summary %>%
  ggplot(., aes(x = specimen_source,
        y = Axis.1)) +
  geom_point(aes(color=specimen_source)) +
  geom_smooth(method = "lm", 
              formula = Axis1 ~ specimen_source, 
              se = F)

# ensure uniform ordering of variables by specimen id
# see lecture 8, slide 37
icu_sorted_summary <- #save
labels(icu_dist) %>%
  enframe(value = "specimen_id") %>%
  select(specimen_id) %>%
  left_join(icu_summary, by = "specimen_id") %>%
  distinct()

# icu distance matrix is used as response variable
adonis(icu_dist ~ specimen_source,
       data = icu_sorted_summary)

```


## Task 4: Dirichlet Multinomial Mixture model

### 4A. DMM model:

Using the Laplace approximation, the best model fit appears to involve 4 Dirichlet mixture components _K_.

```{r task4a, echo=TRUE, eval=TRUE, fig.align='center'}

dmm <- lapply(1:10, # create 10 dmm models
              dmn,
              count = t(icu_mat),
              verbose = FALSE)

lplc <- sapply(dmm, laplace)

ggplot(data = NULL,
         aes(x = seq_along(lplc),
             y = lplc)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(breaks = seq(1,10,2))
  labs(x = "No. of Dirichlet Components - K",
       y = "Model fit (Laplace)")
  
```

### DMM Assignments

```{r task4b, echo=TRUE, eval=TRUE, fig.align='center'}

# select dmm model with best laplace score
best_dmm <- dmm[[which.min(lplc)]]

mixture(best_dmm) %>% 
  as_tibble(rownames = "specimen_id") %>%
  rename_at(.vars = vars(contains("V")),
            .funs = function(x)
              paste0(
                gsub("V","m",x),"_prob")
            ) %>%
  # apply dmm assignment types
  mutate(assignment =
           mixture(best_dmm,
                   assign = TRUE)) ->
  icu_et_dmm_assignments

icu_et_dmm_assignments

# assign categorical mixture type to each specimen
fitted(best_dmm, scale=TRUE) %>% 
  as_tibble(rownames = "otu_id") %>%
  rename_at(.vars = vars(contains("V")),
            .funs = function(x) 
              paste0(gsub("V","m",x),"_fit")) ->
  icu_et_dmm_otu_fits

# plot categorical mixture type by OTU id
icu_et_dmm_otu_fits %>%
  gather(key = which_mix, value = mix_fit, -otu_id) %>%
  ggplot(data = .) +
  geom_tile(mapping = aes(x = which_mix,
                          y = otu_id,
                          fill = mix_fit)) +
  labs(title = "DMM categorical mixture type by OTU id")

```

