---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE)
```

__EPID674: FINAL PROJECT__

* For your final project, you will choose from two data sets to study:

    - (1) an OTU table from a study of respiratory tract microbiome (OP = oropharyngeal, ET = endotracheal), which is described here: https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-016-0151-8   
    - (2) an OTU table from a study of mouse  

* There are four tasks to complete for the final project:  

1. Alpha diversity:  
    - A. Calculate Shannon diversity for all specimens included in your study of choice  
    - B. Plot the Shannon diversity with separate colors for specimen type ("OP" vs "ET" specimens for data set #1; or vendor name for data set #2).  
    - HINT: use the long form of the data and group by specimen  

2. Develop a linear model for the Shannon diversity:  
    - for data set #1: Shannon diversity ~ ET/OP  
    - for data set #2: Shannon diversity ~ vendor 
    - HINT: use the output from above

3. Beta diversity:
    - A. Use the `vegdist` function to calculate Jaccard distances between specimens
    - B. Use the `adonis` function to assess the effect of ET/OP (data set #1) or vendor (data set #2)
    - HINT: use the matrix form of the data... and don't forget to transpose!

4. Dirichlet Multinomial Mixture model:
    - A. Apply a DMM model to your chosen data set -- how many community types are recognized
    - B. Use the "fitted" function to assign categorical mixture type to each specimen
    - HINT: use the matrix form of the data... and don't forget to transpose!



***

__Getting Started: Load the Necesary Packages__

* We have to explicitly load the packages that we will use for our analysis.

```{r load_pkgs, echo=TRUE, eval=TRUE}

# remember to install packages the first time!
library(tidyverse)
library(vegan)
library(ape)
library(DirichletMultinomial)
library(viridis)
set.seed(16)


```

***

__Getting Started: Load the Data__

* Load Data:

```{r load_otus, echo=TRUE, eval=TRUE, fig.align='center'}





```

***

__Task #1__

* Alpha diversity:

```{r task1, echo=TRUE, eval=TRUE, fig.align='center'}





```

***

__Task #2__

* Linear model:

```{r task2, echo=TRUE, eval=TRUE, fig.align='center'}





```

***

__Task #3__

* Beta diversity:

```{r task3, echo=TRUE, eval=TRUE, fig.align='center'}





```

***

__Task #4__

* DMM model:

```{r task4, echo=TRUE, eval=TRUE, fig.align='center'}





```

***

